Case based Arabic Morphological Analysis with Compositional Non-deterministic Automata

### 2010-06-25 tex/sarf.tex
added algorithm and fig

@@

Many morphological analyzers exist~\cite{Sughaiyer:04}. 
Some use specialized part of speech tag sets~\cite{Khoja:01,Darwish:02}.

Buckwalter: used lexicon, added proper nouns, word based, recursive affixes

MAGEAD

Beeseley: many compiled FSAs compared to compositions of three FSAs

SAMA: word based, treebank specific,

MADA+TOKAN: no controller, we do not need a tokenizer

AMIRA: machine learning...

ElixirFM:

\section{Sarf}

@@

\subsection{Motivating Example}
% the FSMs for the three words

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{psfigs/FSM.pstex_t}}
\caption{Example FSM.}
\label{f:examples}
}
\end{figure*}

@@

\subsection{Non-deterministic Composition of FSMs}
% algorithm
% diagram of abstract machines

>>>>

% nondeterministic finite-state automata

@@

The survey work in~\cite{Sughaiyer:04} discusses and compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific specific applications in the morphological 
analyzer itself or use a specific set of part of speech tags
as their reference.
Sarf differs in that it is a general and rich morphological 
analyzer that reports all possible solutions. 
It is case-based in the sense that a controller that drives
the morphological analyzers prunes these decisions. 
Up to our knowledge, Sarf is the only morphological analyzer
that addresses the ``run-on'' words problem.

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs than Buckwalter where it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow recursive 
affixes so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the part of speech information
tags associated with each prefix while these need to be highly 
redundant with Buckwalter and that may lead to consistency issues. 
Sarf keeps the possible analyses alive as a sequence corresponding
to the text. Buckwalter produces a set of solutions for each token 
and the several analysis threads of the whole text need to be 
generated as a product of the token solutions.

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finiste-state automata (FSA). 
Beesley reports that the number of the automata generated by a compiler
from Xerox rules is not controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with a non-deterministic composition of three
FSAs that are the prefix, stem, and suffix FSAs. 
With Sarf it is easy to control the FSAs with a user custom case 
controller, while with Beesley the effort requires recompiling the
Xerox rules.

Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
case study to interfere in judging the analyses much earlier and
at a finer granularity level. 
Also in Sarf there is no need for a separate tokenizer such as
TOKAN as each alive automata carries with it the positions in 
text where a composition decision was made. 
The case-baded controller can build the trace when needed.

%MAGEAD 
    %Maamouri, Mohamed and \bgroup et al.\egroup },

We strongly feel that the refinemet of SAMA~\cite{Maamouri:10} 
is the closest to our approach. SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. We find evidence in this decision that our hypothesis
about the utility of case-based analyzers. The refinements
that resulted in algorithmic changes in SAMA were done manually
and worked in integration with the ATB format only. 
In Sarf we propse a case-based controller approach that can modify 
the behaviour of the analyzer on the fly and that should interact
with the case study. In Sarf the analyzer remains general, and the 
case-based controller specializes the analyses.

The AMIRA~\cite{Diab:07} analyzer uses 
a language independent SVM based analysis. 
While the SVM analysis learns features from the text passed to the case
study, it does not make use of the type oh the query the 
case study targets in the text.

Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
of the Buckwalter analyzer. 
We also use deterministic parsing with tries or our analysis 
to run the prefix, suffix and stem FSAs. 
We think that their inferential-realizational approach 
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can make benefit of many features unique to the Arabic language.
Sarf leaves implementing that to the case-based controller
since in many cases the abstraction needed may not need the 
text to be grammar correct.
We can compare the case-based controller in Sarf to a high level
component in the ElixirFM domain sepcific language.

\section{Sarf}
\label{sec:sarf}

Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with 
proper names and location names extracted from different online 
sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
as well as biblical sources~\cite{}.

\subsection{Sarf example }
\label{sec:example}

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{psfigs/FSM.pstex_t}}
\caption{Example composition of the prefix, stem and suffix FSAs}
\label{f:example}
}
\end{figure*}
% the FSMs for the three words

The compositional FSA diagram in Figure~\ref{f:example}
illustrates the operation of Sarf when parsing the
string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
\RL{n}%
\RL{w}
\RL{b}
\RL{`}
\RL{a} 
\RL{l}
\RL{l}
\RL{a} %
\RL{a}
\RL{h}
\RL{b}
\RL{`}
\RL{l}
\RL{y}
\RL{s}
\RL{w}
in separate form to ease following the example
}
The square and circle legends are states of the FSA
and the directed edges are transitions driven by an input
character that matches the label of the edge. 
The square denotes an accept state, 
the circle denotes a regular state, and the reject states
are omitted for clarity. 
Figure~\ref{f:example}a represents the prefix FSA and 
Figures~\ref{f:example}b and~\ref{f:example}c represent the 
stem and suffix FSAs respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism in this example.

We start at the root square state in the prefix FSA
which is an accept state. 
The edge labeled $\epsilon$ that connects the prefix 
FSA to the stem FSA can transition from any accept state
in the prefix FSA to the root state in the stem FSA.
The edge that connects the stem FSA to the suffix FSA
follows the same behaviour.

When there are two valid transitions such as \RL{w} 
and $\epsilon$ as is the start case, a non-deterministic 
FSA $\Phi$ spawns an exact copy of itself $\Psi$. FSA $\Phi$
makes one transition, and FSA $\Psi$ makes the other. 
Each of the FSAs now represent a valid analysis so far. 
When the FSA is at a state where the next input character
leads to a reject state, the FSA dies.
The reject states are omitted from Figure~\ref{f:example}
and are denoted by the abscence of the corresponding 
characters. 
In our examples, if we assume there were no stems that start 
with the letter \RL{w}, $\Psi$ will die then. 
In reality probably $\Psi$ will die when the input string
is at \RL{wsyl`}.

Note that the prefix and affix FSAs allow recursive 
prefixes. For example, \RL{w} will result in an accept state
that transitions to the stem FSA. 
When \RL{s} follows, we move to another accept state in the 
prefix FSA corresponding to the suffix \RL{ws}. 
The same applies to the suffix FSA.

Lets consider FSA $\Phi$ after it consumed \RL{wsy} 
and transitioned into the stem FSA root node. 
Now $\Phi$ will transition with \RL{l`b} to reach an accept 
state. 
Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
We do that by associating a category value for each accept state. 
We consider the value of the category as part of the state.
Thus each state shown in Figure~\ref{f:example} actually represents
more than one state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in the stem FSA, it will
spawn $\Sigma$ that transitions into the root of the 
suffix FSA. When \RL{h} is consumed, $\Phi$ will die since 
there is no edge from the current state labeled with \RL{h}. 
Now $\Sigma$ represents a valid analysis and can proceed.

Note that $\Sigma$ can report a full word at any accept state
now by spawning a new automata using the $\epsilon$ transition
to the prefix FSA.
This is refined in Sarf to happen only with white space, delimiters, 
or non-connecting characters. 
We left that out in Figure~\ref{f:example} for clarity.

Finally, The case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an automata that 
is not interesting. 
It may also decide to correct the analysis of one automata
based on the decision of the others.

@@

@@

\subsection{Non-deterministic Composition of FSMs}
\label{sec:ndfsa}

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{psfigs/abstract_machine.pstex_t}}
\caption{An abstraction of the non-deterministic composition of the prefix, stem and suffix FSAs}
\label{f:example}
}
\end{figure*}

% algorithm
% diagram of abstract machines

\begin{Verbatim}[fontsize=\relsize{-1},
frame=topline,framesep=4mm,label=\fbox{NDFSA Sarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
Sarf(string $L$, Controller $A$) 
  FSAVec $M$; -- collection of FSA automata
  FSA $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      $A$.control($M$, $c$);
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhiteORNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

      if ( $\Phi$.isWalkable($c$) ) \{
        $\Phi$.transition($c$);
      \} else \{
        $M$.remove($\Phi$);
        $\Phi$.die();
      \} \} \}
\end{Verbatim}
